A recipe for extremely reproducible enrichment analysis

### 2023-04-27 manuscript.Rmd
manuscript updates

<b>Alt titles:</b>

* "Enrichment analysis with gold standard reproducibility"

* "A guide to completely reproducible enrichment analysis"

@@

asddsadasd

@@

Reproducibility is essential to the scientific enterprise by confirming the validity of new discoveries
[1](National Academies of Sciences 2019, p1).
According to a 2016 survey of *Nature* readers, 52% of respondents agreed that there was "significant
crisis" of reproducibility [2](Baker 2016).
While bioinformatics and other branches of computational research are theoretically amenable to
complete reproducibility, in reality this is rarely achieved due to lack of detail provided in methods
sections, lack of shared data and lack of code [3](Peng 2011).
Indeed even if data and code are shared, reproducibility still cannot be guaranteed due to the
differences in computing environments over time, and the availability of dependancies [4](Perkel 2020).

To address this, many articles have been written to endorse the use of "virtual machines" or
"containerisation" to control the "environment" including the operating system and dependancies
along with the research code to facilitate a high degree of reproducibility [4-16](Perkel 2020;
Stein 2010; Dudley & Butte 2010; Sandve et al, 2013; Piccolo & Frampton 2016; Grüning et al, 2014;
Lewis et al, 2016; da Veiga Leprevost et al, 2017; Kulkarni et al, 2018; Pasquier et al, 2018;
Hung et al, 2019; Brito et al, 2020; Piccolo et al, 2021).
Containers also allow for an easy installation process, as all working dependencies are shipped together
and can be installed with a single command.

In addition to shared code, data and containerisation, there are a few ways researchers can make their
work more reproducible.
Firstly, thorough documentation makes the reproducibility process a great deal easier and more accessible
to novice bioinformaticians.
Version control keeps track of code and documentation changes over time and manages the edits made by
multiple contributors.
Literate programming using Rmarkdown or Jupyter notebooks provides optimal documentation by embedding
free text explanations together with the code and outputs such as charts and tables.
Secondly, code and data can be linked. This means that the code knows the location of the the data,
avoiding any "file not found" errors.
Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
for each figure sequentially, making the entire manuscript reproducible by executing just one command
any without any manual steps.
Taken together, computational reproducibility is a spectrum (**Figure 1**), and rests upon five key
pillars mentioned above (**Figure 2**).
Currently, bioinformatics work at the gold standard is rare, despite the availability of such tools for
over a decade.

![Figure 1. The computational research reproducibility spectrum. Creator: Altuna Akalin.](img/repro.webp "The computational research reproducibility spectrum.")

@@

In this article, we focus on enrichment analysis, which is one of the most used methods in
computational biology, and involves the summarisation of omics data to reflect biological changes,
such as the detection of pathway activation in development or disease
[17,18](Slonim 2002; Khatri et al, 2012).
While enrichment analysis is undoubtedly a key method for the interpretation of large
data sets, there are concerns that inadvertant misuse of these methods leads to
unreliable results.
These potential problems include (i) the lack of correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
[19-22](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results.
We are concerned about the reliability of published enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script based methods for this type
of work [21](Wijesooriya et al, 2022).
Although web-based tools are convenient, they are a potential reproducibility concern.
Firstly, incomplete reporting of methods, options and parameters is common.
Secondly, algorithms and functional annotation sets undergo regular updates, and previous versions are
typically not made available hampering later reproducibility.
In some cases the entire web resource can become unavailable, and hence irreproducible due to
a phenomenon known as link decay [23](Hennessey & Ge 2013).

A small number of enrichment analysis applications have been developed to have beneficial properties
of being "containerised" and with archived historical versions [24,25](Ge et al, 2020;
Perampalam & Dick 2020).
This helps reproducibility greatly, but is not a complete solution as it lacks information about how
the input data was used and a record of any parameter selections.

@@

We address this firstly by providing working templates for docker image and Rmarkdown scripts for
common transcriptome and enrichment analysis routines.
Secondly, we provide step-by-step written and video guides to help users customise these templates for
their own work, deploy the analysis, verify the results, and archive the Docker image.

However, before recommending or prescribing such workflows, it is necessary
to assess directly whether current practices are appropriate, by trying to reproduce the enrichment
analyses of previously published studies, and examining whether the conclusions of those articles
are supported or not.
In the following section we investigate the computational reproducibility of enrichment analyses
shown in a sample of 20 journal articles published in 2019.
This should provide a sound basis for recommendations.

@@

From a list of 186 PubMed central articles using enrichment analysis that we examined previously
[5](Wijesooriya et al, 2022), we selected 20 articles that fulfilled the following criteria; (1) the
organism under study is human; (2) the tool used was DAVID Version 6.8 (Database for Annotation,
Visualization, and Integrated Discovery)[101](Huang et al, 2007); (3) the omics type was gene
expression (array or RNA-seq); and (4) that the gene list was provided in the supplementary data.
The basis for these selection criteria is that this combination represents the most common application
of enrichment analysis in our previous study, which could be reproduced without the need to
reanalyse the raw dataset from scratch.

The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
special attention to preventing gene name errors from occurring[102](Zeeberg et al, 2004).
The reproducibility analysis was performed using the respective methodologies stated in the methods
sections of the 20 articles using DAVID version 6.8 (March-May 2022).

@@

### Results and discussion

@@

Taken together, the ability to replicate DAVID enrichment analysis using the supplementary gene
lists was highly variable.
We note that DAVID 6.8 server has been replaced by DAVID 2022 server as of June 1, 2022, meaning
that is no longer possible to reproduce DAVID 6.8 analyses.

@@

The Dockerfile uses a base image from Bioconductor which should correspond to the latest available
stable release version (currently 3.16), running on R v4.2.2 and Ubuntu 22.04.
The image contains instruction for installing a few useful utilities including `nano` and `git` for
modifying scripts, and `magic-wormhole` for transfering data between computers.
It installs R packages from CRAN and Bioconductor repositories to fulfill common tasks in transcriptome
analysis such as fetching RNA-seq counts from DEE2 with `getDEE2` [201](Ziemann et al, 2019),
differential expression with `DESeq2` [202](Love & Huber, 2014),
enrichment analysis with `clusterprofiler`, `fgsea` and `mitch`
[203-205](Yu et al, 2012; Korotkevich et al, 2016; Kaspi & Ziemann 2020).
It also contains instructions to clone a github repository, which contains the scripts that run the
analytical workflow.
Lastly, it copies a gene set library from Reactome [206](Gillespie et al, 2022) with a clear version
history to guarantee provenance of the annotation data.

The Rmarkdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
Rmarkdown allows the unification of extended descriptions, code and results into a single document,
which is becoming ubiquitous in bioinformatics and data science more generally
[207](Grolemund & Wickham 2017, p423).
The ability to provide extended descriptions is helpful for providing contextualising information about
the purpose of the work, detailed methods and interpretations derived from the analysis.
The Rmarkdown script contains code chunks that perform specific tasks including fetching an example
data set from DEE2, inspecting data quality, differential expression and two types of enrichment
analysis; over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 4**).
For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
[22](Zhao & Rhee 2023).

```{r,diagram,echo=FALSE,fig.cap = "Figure 4: Flow chart of example workflow."}

@@

@@

To demonstrate the utility of this workflow, we present the results of this example data analysis;
it is a comparison of control and 5-azacitidine-treated AML3 cell transcriptomes with RNA-seq
in triplicate [208](Lund et al, 2014).
We use this analysis to better understand the effect of DNA methylatransferase inhibitors on
genome regulation and elaborate upon previous findings by Lund et al (2014), which mainly focused on
the relationship between RNA expression and DNA methylation, but did not show results of a
transcriptome-only enrichment analysis.
Furthermore, we provide a comparison of the two most common approaches for enrichment analysis; ORA and
FCS, and give guidance for interpreting results of such tests.

@@

![Figure 5: Differential expression and enrichment analysis with the example workflow.](img/fig_example.png "Example enrichment analysis on")

@@

Although our reproducibility study was relatively small, there are a few key observations.

We observe that only 3/20(15%) of studies achieved a high degree of reproducibility.

A literature search identified very few studies conducted to systematically examine the reproducibility
of bioinformatics research.

There were several case studies, but the overall lack of quantitative data appears to be out of
proportion to the number of review and commentary articles recommending computational best-practices.

Ionnidis et al (2009) examined a set of 10 gene expression studies with available data published in
2005-2006, and found only two could be reproduced to a satisfactory level
(https://www.nature.com/articles/ng.295).

http://www.bioinformatics.deib.polimi.it/geco/workshop/reproducibility.pdf

In the realm of social sciences,

Only 40-50% of social psychology studies could be reproduced, depending on how reproduction is defined
(https://plato.stanford.edu/entries/scientific-reproducibility/)

Reproducibility in the Social Sciences
(https://www.annualreviews.org/doi/abs/10.1146/annurev-soc-090221-035954)

we can conclude that just 15% of studies could be

It is important to note that the DAVID 6.8 server has been replaced by DAVID 2022 server, starting
from June 1, 2022.
For our analysis, DAVID version 6.8 was used for replication procedures.
It was ensured that our analyses got completed by 31 May 2022.
It is no longer possible to reproduce analyses conducted in DAVID 6.8

While there have been many general guides to computatonal reproducibility in the literature,
there is a need for boilerplate code templates that researchers can quickly adapt/tweak for their own
projects.

The demand for this is expected to be huge, as there are massive numbers of published articles featuring
such analyses.
Just in 2022, there were 9,917 PubMed articles with the keywords pathway/enrichment/ontology analysis
in the title or abstract alone.

The more rigourous, well documented and reproducible these studies are, we will find less wasted
research resources and greater trust/support from the general community.

* web tool developers could provide archived versions of the tool as docker images to facilitate
reproducibility while keeping the online version up to date.

As scholars, we should be following best practice in what we do and publish.

Rather than doing the minimum amount of work to get our results published,

Docker install in 32 seconds
Docker pull in 47 seconds

@@

17.  Slonim DK. From patterns to pathways: gene expression data analysis comes of age. Nat Genet [Internet]. 2002 [cited 2023 Mar 3];32 Suppl(S4):502–8. Available from: https://www.nature.com/articles/ng1033z
18.  Khatri P, Sirota M, Butte AJ. Ten years of pathway analysis: current approaches and outstanding challenges. PLoS Comput Biol [Internet]. 2012 [cited 2023 Mar 3];8(2):e1002375. Available from: https://pubmed.ncbi.nlm.nih.gov/22383865/
19.  Timmons JA, Szkop KJ, Gallagher IJ. Multiple sources of bias confound functional enrichment analysis of global -omics data. Genome Biol [Internet]. 2015;16(1):186. Available from: http://dx.doi.org/10.1186/s13059-015-0761-7
20.  Reimand J, Isserlin R, Voisin V, Kucera M, Tannus-Lopes C, Rostamianfar A, et al. Pathway enrichment analysis and visualization of omics data using g:Profiler, GSEA, Cytoscape and EnrichmentMap. Nat Protoc [Internet]. 2019;14(2):482–517. Available from: http://dx.doi.org/10.1038/s41596-018-0103-9
21.  Wijesooriya K, Jadaan SA, Perera KL, Kaur T, Ziemann M. Urgent need for consistent standards in functional enrichment analysis. PLoS Comput Biol [Internet]. 2022;18(3):e1009935. Available from: http://dx.doi.org/10.1371/journal.pcbi.1009935
22.  Zhao K, Rhee SY. Interpreting omics data with pathway enrichment analysis. Trends Genet [Internet]. 2023 [cited 2023 Mar 1];0(0). Available from: https://www.cell.com/trends/genetics/fulltext/S0168-9525(23)00018-5
23.  Hennessey J, Ge S. A cross disciplinary study of link decay and the effectiveness of mitigation techniques. BMC Bioinformatics [Internet]. 2013;14 Suppl 14(S14):S5. Available from: http://dx.doi.org/10.1186/1471-2105-14-S14-S5
24.  Ge SX, Jung D, Yao R. ShinyGO: a graphical gene-set enrichment tool for animals and plants. Bioinformatics [Internet]. 2020 [cited 2023 Mar 9];36(8):2628–9. Available from: https://pubmed.ncbi.nlm.nih.gov/31882993/
25.  Perampalam P, Dick FA. BEAVR: a browser-based tool for the exploration and visualization of RNA-seq data. BMC Bioinformatics [Internet]. 2020 [cited 2023 Mar 9];21(1):221. Available from: https://pubmed.ncbi.nlm.nih.gov/32471392/

101.  Huang DW, Sherman BT, Tan Q, Collins JR, Alvord WG, Roayaei J, et al. The DAVID Gene Functional Classification Tool: a novel biological module-centric algorithm to functionally analyze large gene lists. Genome Biol [Internet]. 2007;8(9):R183. Available from: http://dx.doi.org/10.1186/gb-2007-8-9-r183
102.  Zeeberg BR, Riss J, Kane DW, Bussey KJ, Uchio E, Linehan WM, et al. Mistaken identifiers: gene name errors can be introduced inadvertently when using Excel in bioinformatics. BMC Bioinformatics [Internet]. 2004;5:80. Available from: http://dx.doi.org/10.1186/1471-2105-5-80

201.  Ziemann M, Kaspi A, El-Osta A. Digital expression explorer 2: a repository of uniformly processed RNA sequencing data. Gigascience [Internet]. 2019;8(4). Available from: http://dx.doi.org/10.1093/gigascience/giz022
202.  Love MI, Huber W, Anders S. Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biol [Internet]. 2014;15(12):550. Available from: http://dx.doi.org/10.1186/s13059-014-0550-8
203.  Yu G, Wang LG, Han Y, He QY. clusterProfiler: an R package for comparing biological themes among gene clusters. OMICS [Internet]. 2012;16(5):284–7. Available from: http://dx.doi.org/10.1089/omi.2011.0118
204.  Korotkevich G, Sukhov V, Budin N, Shpak B, Artyomov MN, Sergushichev A. Fast gene set enrichment analysis [Internet]. bioRxiv. 2016. p. 060012. Available from: http://biorxiv.org/content/early/2021/02/01/060012.abstract
205.  Kaspi A, Ziemann M. Mitch: Multi-contrast pathway enrichment for multi-omics and single-cell profiling data. BMC Genomics [Internet]. 2020;21(1):447. Available from: http://dx.doi.org/10.1186/s12864-020-06856-9
206.  Gillespie M, Jassal B, Stephan R, Milacic M, Rothfels K, Senff-Ribeiro A, et al. The reactome pathway knowledgebase 2022. Nucleic Acids Res [Internet]. 2022;50(D1):D687–92. Available from: http://dx.doi.org/10.1093/nar/gkab1028
207.  Grolemund G, Wickham H. R for Data Science. Sebastopol, CA: O’Reilly Media; 2017.
208.  Lund K, Cole JJ, VanderKraats ND, McBryan T, Pchelintsev NA, Clark W, et al. DNMT inhibitors reverse a specific signature of aberrant promoter DNA methylation and associated gene silencing in AML. Genome Biol [Internet]. 2014;15(8):406. Available from: http://dx.doi.org/10.1186/s13059-014-0406-2

## Supplementary information

Table XX: The software bundles in the example image.
R package dependancies not shown.

>>>>

@@

Irreproducible research is a significant problem that wastes resources undermines the public
perception of science.
Previous work has highlighted that published enrichment analyses suffers from statistical and reporting
flaws.
We sought to determine whether this translates into irreproducibility by examining whether the
findings of 20 open-access articles describing enrichment analysis were reproducible.
We find that only three articles exhibited a high degree of concordance, while seven exhibited poor
reproducibility.
The findings indicate that current enrichment analysis practices are inadequate.
To address this, we provide a set of boilerplate scripts and a step-by-step protocol to adapt them to
a new project's needs.
This recipe is built according to a framework we call "five pillars of computational reproducibility"
, which when implemented together facilitate extreme reproducibility.

@@

Reproducibility is essential to the scientific enterprise by confirming the validity of new discoveries
[1](National Academies of Sciences 2019, p1).
According to a 2016 survey of *Nature* readers, 52% of respondents agreed that there was a "significant
crisis" of reproducibility [2](Baker 2016).
While bioinformatics and other branches of computational research are theoretically amenable to
complete reproducibility, in reality this is rarely achieved due to lack of detail provided in methods
sections, lack of shared data and lack of code [3](Peng 2011).
Indeed even if data and code are shared, reproducibility still cannot be guaranteed due to the
differences in computing environments over time, and the availability of dependancies [4](Perkel 2020).

To address this, many articles have been written to endorse the use of "virtual machines" or
"containerisation" to control the computing environment including the operating system and dependancies
along with the research code to facilitate a high degree of reproducibility [4-16](Perkel 2020;
Stein 2010; Dudley & Butte 2010; Sandve et al, 2013; Piccolo & Frampton 2016; Grüning et al, 2014;
Lewis et al, 2016; da Veiga Leprevost et al, 2017; Kulkarni et al, 2018; Pasquier et al, 2018;
Hung et al, 2019; Brito et al, 2020; Piccolo et al, 2021).
Containers also allow for an easy installation process, as all working dependencies are shipped together
and can be installed with a single command.

In addition to shared code, data and containerisation, there are a few ways researchers can make their
work more reproducible.
Firstly, thorough documentation makes the reproducibility process a great deal easier and more accessible
to novice bioinformaticians.
Version control keeps track of code and documentation changes over time and manages the edits made by
multiple contributors.
Literate programming using Rmarkdown or Jupyter notebooks provides optimal documentation by embedding
free text explanations together with the code and outputs such as charts and tables.
Secondly, code and data can be linked. This means that the code knows the location of the the data,
avoiding any "file not found" errors.
Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
for each figure sequentially, making the entire manuscript reproducible by executing just one command
and without any manual steps.
Taken together, computational reproducibility is a spectrum (**Figure 1**) [17](Akalin 2018), and rests upon five key
pillars mentioned above (**Figure 2**).
Currently, bioinformatics work at the gold standard is rare, despite the availability of such tools for
over a decade.

![Figure 1. The computational research reproducibility spectrum. Creator: Altuna Akalin. Republished with permission [17](Akalin 2018).](img/repro.webp "The computational research reproducibility spectrum.")

@@

In this article, we focus on enrichment analysis, which is one of the most used methods in
computational biology, and involves the summarisation of omics data to reflect biological changes,
such as the detection of pathway activation in development or disease
[18,19](Slonim 2002; Khatri et al, 2012).
While enrichment analysis is undoubtedly a key method for the interpretation of large
data sets, there are concerns that inadvertant misuse of these methods leads to
unreliable results.
These potential problems include (i) the lack of correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
[20-23](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results.
We are concerned about the reliability of published enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script based methods for this type
of work [22](Wijesooriya et al, 2022).
Although web-based tools are convenient, they are a potential reproducibility concern.
Firstly, incomplete reporting of methods, options and parameters is common.
Secondly, algorithms and functional annotation sets undergo regular updates, and previous versions are
typically not made available, hampering later reproducibility.
In some cases the entire web resource can become unavailable, and hence irreproducible due to
a phenomenon known as link decay [24](Hennessey & Ge 2013).

A small number of enrichment analysis applications have been developed to have beneficial properties
of being "containerised" and with archived historical versions [25,26](Ge et al, 2020;
Perampalam & Dick 2020).
This helps reproducibility greatly, but is not a complete solution as it lacks information about how
the input data was used and a record of any parameter selections.

@@

We address this firstly by providing working templates for Docker image and Rmarkdown scripts for
common transcriptome and enrichment analysis routines.
Secondly, we provide step-by-step written and video guides to help users customise these templates for
their own work, deploy the analysis, verify the results, and share the Docker image.

However, before recommending or prescribing such workflows, it is necessary
to assess directly whether current practices are adequate, by trying to reproduce the enrichment
analyses of previously published studies, and examining whether the conclusions of those articles
are supported or not.
In the following section we investigate the computational reproducibility of enrichment analyses
shown in a sample of 20 open-access journal articles published in 2019.
This should provide a sound basis for recommendations.

@@

From a list of 186 PubMed central articles using enrichment analysis that we examined previously
[22](Wijesooriya et al, 2022), we selected 20 articles that fulfilled the following criteria; (1) the
organism under study is human; (2) the tool used was DAVID Version 6.8 (Database for Annotation,
Visualization, and Integrated Discovery)[27](Huang et al, 2007); (3) the omics type was gene
expression (array or RNA-seq); and (4) that the gene list was provided in the supplementary data.
The basis for these selection criteria is that this combination represents the most common application
of enrichment analysis in our previous study, which could be reproduced without the need to
reanalyse the raw dataset from scratch. These articles are listed in the bibliography [28-47]().

The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
special attention to preventing gene name errors from occurring [48](Zeeberg et al, 2004).
The reproducibility analysis was performed using the respective methodologies stated in the methods
sections of the 20 articles using DAVID version 6.8 (March-May 2022).

@@

### Results

@@

Taken together, the ability to replicate DAVID enrichment analysis using the supplementary gene
lists was highly variable.

@@

The Dockerfile uses a base image from Bioconductor which should correspond to the latest available
stable release version (currently 3.16), running on R v4.2.2 and Ubuntu 22.04.
The image contains instruction for installing a few useful utilities including `nano` and `git` for
modifying scripts, and `magic-wormhole` for transfering data between computers.
It installs R packages from CRAN and Bioconductor repositories to fulfill common tasks in transcriptome
analysis such as fetching RNA-seq counts from DEE2 with `getDEE2` [49](Ziemann et al, 2019),
differential expression with `DESeq2` [50](Love & Huber, 2014),
enrichment analysis with `clusterprofiler`, `fgsea` and `mitch`
[51-53](Yu et al, 2012; Korotkevich et al, 2016; Kaspi & Ziemann 2020).
It also contains instructions to clone a github repository, which contains the scripts that run the
analytical workflow.
Lastly, it copies a gene set library from Reactome [54](Gillespie et al, 2022) with a clear version
history to guarantee provenance of the annotation data.

The Rmarkdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
Rmarkdown allows the unification of extended descriptions, code and results into a single document,
which is becoming ubiquitous in bioinformatics and data science more generally
[55](Grolemund & Wickham 2017, p423).
The ability to provide extended descriptions is helpful for providing contextualising information about
the purpose of the work, detailed methods and interpretations derived from the analysis.
The Rmarkdown script contains code chunks that perform specific tasks including fetching an example
data set from DEE2, inspecting data quality, differential expression and two types of enrichment
analysis; over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 4**).
For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
[23](Zhao & Rhee 2023).

```{r,diagram,echo=FALSE,fig.cap = "Figure 4. Flow chart of example workflow."}

@@

The software versions used in this example analysis are provided in Table 2.

: Table 2. The software bundled in the example Docker image.

@@

To demonstrate the utility of this workflow, we present the results of this example data analysis;
it is a comparison of control and 5-azacitidine-treated AML3 cell transcriptomes with RNA-seq
in triplicate [56](Lund et al, 2014).
We use this analysis to better understand the effect of DNA methylatransferase inhibitors on
genome regulation and elaborate upon previous findings by Lund et al (2014), which mainly focused on
the relationship between RNA expression and DNA methylation, but did not show results of a
transcriptome-only enrichment analysis.
Furthermore, we provide a comparison of the two most common approaches for enrichment analysis; ORA and
FCS, and give guidance for interpreting results of such tests.

@@

![Figure 5. Differential expression and enrichment analysis with the example workflow.](img/fig_example.png "Example enrichment analysis on")

@@

Although our reproducibility study was relatively small, there are a few key observations.
We observe that only 3/20(15%) of studies achieved a high degree of reproducibility.
In a way, this is surprising, given the simplicity of using DAVID, only
requiring the user to provide a list of genes in a text box and select some options.
These irregularities may have arisen inadvertently from several sources:

* The methods stated in the respective articles are inconsistent with what was actually done.

* The list of genes in the supplement is not the same as what was used for enrichment analysis.

* Ambiguity around whether up- and down-regulate genes were considered
separately or together

* Ambiguity around whether all GO classes were considered or just one, such as "biological process".

* Ambiguity around whether a custom background gene set was used.

The observed lack of reproducibility might be expected given that substantial methodological flaws
were present in 85% of enrichment analyses [22](Wijesooriya et al, 2022).
A non-exhaustive literature search identified a small number of case studies, but a dearth of systematic
examinations of bioinformatics research reproducibility.
Ionnidis et al (2009)[57] examined a set of 10 gene expression studies with available data published in
2005-2006, and found only two could be reproduced to a satisfactory level.

This problem is not limited to bioinformatics.
In the field of drug development, according to a 2011 study by Bayer, only 25% of projects could be
replicated to a high degree [58](Prinz et al, 2011).
When the US Biotech firm Amgen set out to reproduce 53 landmark cancer studies, they were able to
reproduce only 6 (11%) [59](Begley & Ellis 2012).

It is clear that as a field, the life sciences need to raise the rate of reproducibility or risk the
public perceiving it as unreliable and wasteful.

We propose the five pillars (**Figure 2**), as a framework to guarantee computational reproducibility.
However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
protocol to guide them on how to customise boilerplate Dockerfile and Rmarkdown scripts to suit their
project's needs.

Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
we estimate at 15-20 hours.
This is significantly longer than using a web-based tool that provides results in just a few
minutes, but we argue the time investment is well spent.
It ensures analytical reproducibility, assists with auditing and makes the process of reproduction
simple and fast (Box 1), which should aid the early identification of errors.

We also want to emphasise that there is still a definite need for graphical and web-based tools.
These are good for data exploration, to quickly identify trends that can be followed up with more
rigourous methods.
Moreover, a lot of work is being done to make such point-and-click tools more reproducible.
The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
workflows that enable reproducibility [60-62](Kuehn et al; 2008; Goecks et al, 2010; Teichman et al, 2023).
The Degust differential expression suite is a graphical interface to popular
expression tools like edgeR and limma, and allows users to download the script to facilitate future
non-interactive reproducibility [63](Powell 2019).
Such functionality could be written into existing enrichment analysis tools like DAVID, and together with
retention of historical gene annotation sets could provide a "good enough" solution to reproducibility.

Although we hope that extremely reproducible research will become more widespread, we do not
anticipate any dramatic changes in researcher behaviour unless the motivating factors are augmented.
Reward structures in academia encourage individuals to publish many articles, and reproducibility
is not a significant factor unless the issues are extreme enough to provoke a retraction by the journal,
which could tarnish the reputation of an author or their institution.
Solutions to this could include mandates by funding agencies or institutions [64](Garijo et al, 2013),
and/or badging of articles that comply with best practices [65](Blohowiak et al, 2023).

## Availability of materials

* Protocol: protocols.io

* Boilerplate code: https://github.com/markziemann/enrichment_recipe

* Example Docker image: https://hub.docker.com/r/mziemann/enrichment_recipe

@@

17.  Akalin A. Scientific data analysis pipelines and reproducibility. Towards Data Science. 2018. Available from: https://towardsdatascience.com/scientific-data-analysis-pipelines-and-reproducibility-75ff9df5b4c5
18.  Slonim DK. From patterns to pathways: gene expression data analysis comes of age. Nat Genet [Internet]. 2002 [cited 2023 Mar 3];32 Suppl(S4):502–8. Available from: https://www.nature.com/articles/ng1033z
19.  Khatri P, Sirota M, Butte AJ. Ten years of pathway analysis: current approaches and outstanding challenges. PLoS Comput Biol [Internet]. 2012 [cited 2023 Mar 3];8(2):e1002375. Available from: https://pubmed.ncbi.nlm.nih.gov/22383865/
20.  Timmons JA, Szkop KJ, Gallagher IJ. Multiple sources of bias confound functional enrichment analysis of global -omics data. Genome Biol [Internet]. 2015;16(1):186. Available from: http://dx.doi.org/10.1186/s13059-015-0761-7
21.  Reimand J, Isserlin R, Voisin V, Kucera M, Tannus-Lopes C, Rostamianfar A, et al. Pathway enrichment analysis and visualization of omics data using g:Profiler, GSEA, Cytoscape and EnrichmentMap. Nat Protoc [Internet]. 2019;14(2):482–517. Available from: http://dx.doi.org/10.1038/s41596-018-0103-9
22.  Wijesooriya K, Jadaan SA, Perera KL, Kaur T, Ziemann M. Urgent need for consistent standards in functional enrichment analysis. PLoS Comput Biol [Internet]. 2022;18(3):e1009935. Available from: http://dx.doi.org/10.1371/journal.pcbi.1009935
23.  Zhao K, Rhee SY. Interpreting omics data with pathway enrichment analysis. Trends Genet [Internet]. 2023 [cited 2023 Mar 1];0(0). Available from: https://www.cell.com/trends/genetics/fulltext/S0168-9525(23)00018-5
24.  Hennessey J, Ge S. A cross disciplinary study of link decay and the effectiveness of mitigation techniques. BMC Bioinformatics [Internet]. 2013;14 Suppl 14(S14):S5. Available from: http://dx.doi.org/10.1186/1471-2105-14-S14-S5
25.  Ge SX, Jung D, Yao R. ShinyGO: a graphical gene-set enrichment tool for animals and plants. Bioinformatics [Internet]. 2020 [cited 2023 Mar 9];36(8):2628–9. Available from: https://pubmed.ncbi.nlm.nih.gov/31882993/
26.  Perampalam P, Dick FA. BEAVR: a browser-based tool for the exploration and visualization of RNA-seq data. BMC Bioinformatics [Internet]. 2020 [cited 2023 Mar 9];21(1):221. Available from: https://pubmed.ncbi.nlm.nih.gov/32471392/
27.  Huang DW, Sherman BT, Tan Q, Collins JR, Alvord WG, Roayaei J, et al. The DAVID Gene Functional Classification Tool: a novel biological module-centric algorithm to functionally analyze large gene lists. Genome Biol [Internet]. 2007;8(9):R183. Available from: http://dx.doi.org/10.1186/gb-2007-8-9-r183
28.  Liu H, Wu N, Zhang Z, Zhong X, Zhang H, Guo H, Nie Y, Liu Y. Long Non-coding RNA LINC00941 as a Potential Biomarker Promotes the Proliferation and Metastasis of Gastric Cancer. Front Genet. 2019 Jan 22;10:5. doi: 10.3389/fgene.2019.00005.
29.  Zhang S, Cao R, Li Q, Yao M, Chen Y, Zhou H. Comprehensive analysis of lncRNA-associated competing endogenous RNA network in tongue squamous cell carcinoma. PeerJ. 2019 Feb 6;7:e6397. doi: 10.7717/peerj.6397.
30.  Zhang L, Luo M, Yang H, Zhu S, Cheng X, Qing C. Next-generation sequencing-based genomic profiling analysis reveals novel mutations for clinical diagnosis in Chinese primary epithelial ovarian cancer patients. J Ovarian Res. 2019 Feb 20;12(1):19. doi: 10.1186/s13048-019-0494-4.
31.  Wang K, Li L, Fu L, Yuan Y, Dai H, Zhu T, Zhou Y, Yuan F. Integrated Bioinformatics Analysis the Function of RNA Binding Proteins (RBPs) and Their Prognostic Value in Breast Cancer. Front Pharmacol. 2019 Mar 1;10:140. doi: 10.3389/fphar.2019.00140.
32.  Balusamy SR, Ramani S, Natarajan S, Kim YJ, Perumalsamy H. Integrated transcriptome and in vitro analysis revealed anti-proliferative effect of citral in human stomach cancer through apoptosis. Sci Rep. 2019 Mar 19;9(1):4883. doi: 10.1038/s41598-019-41406-8.
33.  Ahn KS, Kang KJ, Kim YH, Kim TS, Song BI, Kim HW, O'Brien D, Roberts LR, Lee JW, Won KS. Genetic features associated with 18F-FDG uptake in intrahepatic cholangiocarcinoma. Ann Surg Treat Res. 2019 Apr;96(4):153-161. doi: 10.4174/astr.2019.96.4.153.
34.  Chai YJ, Chae H, Kim K, Lee H, Choi S, Lee KE, Kim SW. Comparative Gene Expression Profiles in Parathyroid Adenoma and Normal Parathyroid Tissue. J Clin Med. 2019 Mar 2;8(3):297. doi: 10.3390/jcm8030297.
35.  Guo C, Xu LF, Li HM, Wang W, Guo JH, Jia MQ, Jia R, Jia J. Transcriptomic study of the mechanism of anoikis resistance in head and neck squamous carcinoma. PeerJ. 2019 May 23;7:e6978. doi: 10.7717/peerj.6978.
36.  Alebrahim D, Nayak M, Ward A, Ursomanno P, Shams R, Corsica A, Sleiman R, Fils KH, Silvestro M, Boytard L, Hadi T, Gelb B, Ramkhelawon B. Mapping Semaphorins and Netrins in the Pathogenesis of Human Thoracic Aortic Aneurysms. Int J Mol Sci. 2019 Apr 28;20(9):2100. doi: 10.3390/ijms20092100.
37.  Hong H, Liu T, Wu H, Zhang J, Shi X, Le X, Chen AM, Mo H, Huang Q, Zhou H, Rao X. MicroRNA-550a is associated with muscle system conferring poorer survival for esophageal cancer. Biosci Rep. 2019 May 31;39(5):BSR20181173. doi: 10.1042/BSR20181173.
38.  Han MZ, Wang S, Zhao WB, Ni SL, Yang N, Kong Y, Huang B, Chen AJ, Li XG, Wang J, Wang DH. Immune checkpoint molecule herpes virus entry mediator is overexpressed and associated with poor prognosis in human glioblastoma. EBioMedicine. 2019 May;43:159-170. doi: 10.1016/j.ebiom.2019.04.002.
39.  McSweeney KM, Bozza WP, Alterovitz WL, Zhang B. Transcriptomic profiling reveals p53 as a key regulator of doxorubicin-induced cardiotoxicity. Cell Death Discov. 2019 Jun 12;5:102. doi: 10.1038/s41420-019-0182-6.
40.  Wang C, Li D, Zhang L, Jiang S, Liang J, Narita Y, Hou I, Zhong Q, Zheng Z, Xiao H, Gewurz BE, Teng M, Zhao B. RNA Sequencing Analyses of Gene Expression during Epstein-Barr Virus Infection of Primary B Lymphocytes. J Virol. 2019 Jun 14;93(13):e00226-19. doi: 10.1128/JVI.00226-19.
41.  Pan Q, Wang L, Zhang H, Liang C, Li B. Identification of a 5-Gene Signature Predicting Progression and Prognosis of Clear Cell Renal Cell Carcinoma. Med Sci Monit. 2019 Jun 13;25:4401-4413. doi: 10.12659/MSM.917399.
42.  Wang Y, Zhao W, Liu X, Guan G, Zhuang M. ARL3 is downregulated and acts as a prognostic biomarker in glioma. J Transl Med. 2019 Jun 24;17(1):210. doi: 10.1186/s12967-019-1914-3.
